Relational Parametricity for Algebraically Indexed Types

### 2012-05-29 paper/section-general-framework.tex
Add the index-erasure semantics of programs, and tidy up the proof of the abstraction theorem

The well-typed programs are defined in
\autoref{fig:programs}. \textbf{FIXME: define the erasure
  interpretation of terms}.

@@

% For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
% index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
% structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
% \ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
% one-element set.

\begin{theorem}[Parametricity]
  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
\end{theorem}

\begin{proof}
  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
  three interesting cases follow:
  \begin{description}
  \item[Case \TirName{TyEq}] This case follows directly from
    \lemref{lem:tyeqsubst-relational}.
  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
    $A = \forall i\mathord:s. A'$. We know from the induction
    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho'}$,
    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{\rho'}$. We are given $\rho \in
    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
    \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho''}$ by \lemref{lem:tysubst-rel}. So we
    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}{\rho''}$, and thus
    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
      i.M'}\eta_2) \in \rsem{\forall
      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
    know that for all $\rho' \in \extends{\rho}{i:s}$,
    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{\rho'}$. If we let $\rho' = \rho \circ
    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
    as required.
  \end{description}
  The remaining cases are standard for proofs of the fundamental lemma
  of logical relations for simply-typed systems.
\end{proof}

>>>>

The well-typed programs are defined in \autoref{fig:programs}. FIXME:
explain each of the rules.

@@

\subsubsection{The Index-Erasure Interpretation of Programs}
\label{sec:erasure-semantics-programs}

For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
one-element set.

% FIXME: do we need to prove something about the effect
% of substitution on the erasure semantics of contexts?

For a well-typed program $\Delta; \Gamma \vdash M : A$, we define the
\emph{erasure interpretation} as a function $\tmSem{e} :
\ctxtSem{\Gamma} \to \tySem{A}$, completely ignoring the indexing
information. In light of \lemref{lem:tyeqsubst-erasure}, we define
this function directly on the syntax of well-typed terms, rather than
on the typing derivations. We define $\tmSem{e}$ by the following
clauses:
\begin{eqnarray*}
  \tmSem{x}\eta & = & \eta_x \\
  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
  \left\llbracket
    \begin{array}{l}
      \textrm{case}\ M\ \textrm{of}\\
      \textrm{inl}\ x.N_1;\\
      \textrm{inr}\ y.N_2
    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
  \left\{
    \begin{array}{ll}
      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
    \end{array}
  \right. \\
  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
\end{eqnarray*}
In the above, we have written $\eta_x$ to denote the appropriate
projection from $\eta$ to get the position corresponding to the
variable $x$ in the context.

\subsubsection{The Abstraction Theorem}
\label{sec:abstraction-theorem}

We now state and prove the abstraction theorem for well-typed programs
in our system. The abstraction theorem states that the type erasure
interpretation of a well-typed program is related to itself by the
relational interpretation of its type. From this result we will be
able to deduce interesting properties of programs adhering to the
algebraically-indexed typing discipline in
\autoref{sec:instantiations}. % FIXME: spruce this up when the instantiations section is actually written.

To state and prove the abstraction theorem for open terms, we first
need to extend the relational interpretation of types to typing
contexts. The relational interpretation of contexts is defined by the
following clauses, where we have used the pairing operation
$-\relTimes-$ defined in \autoref{sec:relational-semantics}.
\begin{eqnarray*}
  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \{(*,*)\} \\
  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
\end{eqnarray*}
The relational interpretation of contexts inherits from the relational
interpretation of types the property of interpretating the application
of simultaneous substitutions as composition:
\begin{lemma}\label{lem:ctxtsubst-rel}
  If $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
\end{lemma}

\begin{theorem}[Abstraction]
  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
\end{theorem}

\begin{proof}
  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
  three interesting cases follow:
  \begin{description}
  \item[Case \TirName{TyEq}] This case follows directly from
    \lemref{lem:tyeqsubst-relational}.
  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
    $A = \forall i\mathord:s. A'$. We know from the induction
    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho'}$,
    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{\rho'}$. We are given $\rho \in
    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
    \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho''}$ by \lemref{lem:ctxtsubst-rel}. So we
    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}{\rho''}$, and thus
    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
      i.M'}\eta_2) \in \rsem{\forall
      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
    know that for all $\rho' \in \extends{\rho}{i:s}$,
    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{\rho'}$. If we let $\rho' = \rho \circ
    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
    by \lemref{lem:ctxtsubst-rel}, conclude that $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
    as required.
  \end{description}
  The remaining cases are standard for proofs of the fundamental lemma
  of logical relations for simply-typed systems.
\end{proof}
